Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-27 paper.tex
A few updates

\author[mcgill]{Reza Adalat}
\author[mcgill]{Natacha Beck}
\author[mcgill]{Samir Das}
\author[isi]{Rafael Ferreira da Silva}
\author[mcgill]{Najmeh Khalili-Mahani}
\author[criugm]{Pierre-Olivier Quirion}
\author[mcgill]{Pierre Rioux}

@@

\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montreal, QC, Canada.}
\address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways. Based on our experience with
  the development and operations of the CBRAIN and VIP science
  gateways, and our past involvement in the SHIWA project, we abstract
  software components and interactions in 6 architectures that we use
  to classify 18 systems. We compare these architectures using metrics
  measuring global complexity, robustness, extensibility, scalability
  and other specific features. Results show the pros and cons of the
  different architectures: tight integration and sub-tasking are the
  simplest architectures and should therefore be the easiest to
  implement and the most robust ones, most architectures perform the
  same in terms of extensibility, the pool model is the most scalable
  and meta-workflows are only available in nested workflows and
  workflow import. These results provide insights for software
  architects on how to integrate science gateways and workflow
  engines.
\end{abstract}

\begin{keyword}
Workflow engines \sep science gateways \sep software architecture.
\end{keyword}

@@

Workflow engines are critical components of the ecosystem of tools and
services offered by science gateways to leverage distributed
infrastructures for the efficient execution of applications through
the web. Several software architectures can be adopted to integrate
workflow engines in science gateways, with important consequences on
the required development effort and on the resulting system. This
paper describes, exemplifies and compares such architectures, based on
system-independent representations of their main components and
interactions.

@@

In the last decade, the e-Science community has developed workflow
systems to help developers access distributed infrastructures such as
clusters, grids, clouds and web services. These efforts resulted in
tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and others. Such workflow engines usually
describe applications in a high-level language that offers specific
data and control flow constructs, parallelization operators, visual
edition features, links with domain-specific tool repositories,
provenance recording, among other features. An overview of workflow
systems features and capability is available
in~\cite{deelman2009workflows}.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to handle high-throughput
task executions. For instance, FSL can launch tasks on SGE through its
fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These
domain-specific engines represent a tremendous opportunity for science
gateways to leverage existing tools and applications. They nicely
complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

@@

Science gateways are used to share resources within a community and to
provide increased performance and capacity through facilitated access
to storage and computing power. They are often accessible through a
web interface that helps users manage access rights, data transfers,
task execution and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications.

@@

The architectures described hereafter also build on experience with
the SHIWA Simulation Platform~\cite{terstyanszky2014enabling}. The
SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
developed in the European projects
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}. It focuses on workflow
interoperability for various disciplines, in particular through the
building of meta-workflows, i.e. workflows that are built out of other
workflows. It implements the so-called Coarse-Grained Interoperability
framework where workflow engines are integrated in the science gateway
as other applications and invoked through a dedicated service. The
Fine-Grained Interoperability framework was also developed in the
SHIWA project, where conversions between workflow languages are
achieved through a common intermediate
representation~\cite{plankensteiner-prodan-etal:2013}. \note{Tristan}{Silvia,
  do you think it is legitimate to refer to SHIWA like this in the
  paper, or should we involve Gabor in the writing?}

@@

The interactions involved in the architectures are described below and
labeled consistently with the notations used on
Figure~\ref{fig:architectures}.

@@

  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}]{ \def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (sub-tasking).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler. 8': (PSOM only) PSOM master and workers
  execute workflow. 9. Execution server transfers results to storage
  server(s). Interaction \texttt{b} in Figure~\ref{fig:architectures}
  is implemented by steps \texttt{4}, \texttt{5}, \texttt{7} and
  \texttt{8} (regular interactions with batch manager). Interaction
  \texttt{c} is implemented by steps \texttt{3} and
  \texttt{9}. Interaction \texttt{e} is implemented by step \texttt{6}
  (for FSL: through a modified version of the fsl\_sub script
  available in \url{https://github.com/aces/cbrain-plugins-neuro}; for
  PSOM: through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

>>>>

\author[mcgill]{Reza Adalat}
\author[mcgill]{Natacha Beck}
\author[mcgill]{\\Samir Das}
\author[isi]{Rafael Ferreira da Silva}
\author[mcgill]{Najmeh Khalili-Mahani}
\author[criugm]{Pierre-Olivier Quirion}
\author[mcgill]{Pierre Rioux}

@@

\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montr\'eal, QC, Canada.}
\address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways, based on our experience with
  the CBRAIN and VIP platforms, and on our past involvement in the
  SHIWA project. We abstract software components and interactions in 6
  architectures that we use to classify 18 systems. We compare these
  architectures using metrics measuring global complexity, robustness,
  extensibility, scalability and other specific features. Results show
  the pros and cons of the different architectures: tight integration
  and sub-tasking are the simplest architectures and should therefore
  be the easiest to implement and the most robust ones, most
  architectures perform the same in terms of extensibility, the pool
  model is the most scalable and meta-workflows are only available in
  nested workflows and workflow import. Our analysis provides insights
  for software architects on how to integrate science gateways and
  workflow engines.
\end{abstract}

\begin{keyword}
Workflow engines \sep science gateways \sep software architectures.
\end{keyword}

@@

Workflow engines are critical components in the ecosystem of tools and
services offered by science gateways for the efficient and transparent
execution of applications on distributed infrastructures. Several
software architectures can be adopted to integrate workflow engines in
science gateways, with important consequences on the required
development effort and on the resulting system. This paper describes,
exemplifies and compares such architectures, based on
system-independent representations of their main components and
interactions.

@@

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools such as
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and others. Such workflow engines usually
describe applications in a high-level language with specific data and
control flow constructs, parallelization operators, visual edition
features, links with domain-specific tool repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular clusters, to execute tasks at
a high throughput. For instance, FSL can launch tasks on SGE through
its fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

@@

Science gateways are used to share resources within a community and to
provide increased performance and capacity through facilitated access
to storage and computing power. They are often accessible through a
web interface that helps users manage access rights, data transfers,
task execution, and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications.

@@

The architectures described hereafter also build on experience with
the SHIWA Simulation Platform~\cite{terstyanszky2014enabling}. The
SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
developed in the European projects
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}. It focuses on workflow
interoperability for various disciplines, in particular through the
building of meta-workflows, i.e. workflows that are built out of other
workflows. It implements the so-called Coarse-Grained Interoperability
framework where workflow engines are integrated in the science gateway
as other applications and invoked through a dedicated service. The
Fine-Grained Interoperability framework was also developed in the
SHIWA project, where conversions between workflow languages are
achieved through a common intermediate
representation~\cite{plankensteiner-prodan-etal:2013}.

@@

The interactions involved in the architectures are described below and
labeled consistently with Figure~\ref{fig:architectures}.

@@

  with heterogeneous interfaces. \closedanswerednote{Marc-Etienne}{this -c2- can be also trigerred

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}]{ \def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (sub-tasking).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler. FSL sub-tasks will run locally
  instead of being submitted to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures} is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}
